---
title: "11636_Data analysis"
author: "Filip Noijons"
date: "`r format(Sys.Date())`"
output:
  github_document:
    toc: true
    toc_depth: 3
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile,
                          encoding = encoding,
                          output_file = "README.md")
      })
params:
  tumor_file: "~/export_catrin/11636_tidy_tumor_volumes_filtered.xlsx"
  bw_file: "~/export_catrin/11636_bodyweight_mouse_and_group_means.xlsx"
  outdir: "~/export_catrin/11636"
  bin_width: 3
  bin_anchor: 1
  bin_label: "start"
  min_n_bodyweight: 5
---



# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Packages
required_pkgs <- c("readxl", "dplyr", "ggplot2", "forcats", "knitr")
to_install <- setdiff(required_pkgs, rownames(installed.packages()))
if (length(to_install)) install.packages(to_install, dependencies = TRUE, repos = "https://cloud.r-project.org")

library(readxl)
library(dplyr)
library(ggplot2)
library(forcats)
library(knitr)
library(purrr)

# Params
tumor_file <- path.expand(params$tumor_file)
bw_file    <- path.expand(params$bw_file)
outdir     <- params$outdir

bin_width  <- as.integer(params$bin_width)
bin_anchor <- as.integer(params$bin_anchor)
bin_label  <- as.character(params$bin_label)
if (!bin_label %in% c("start","center")) bin_label <- "start"
minN_bw    <- as.integer(params$min_n_bodyweight)

if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE, showWarnings = FALSE)
stopifnot(file.exists(tumor_file), file.exists(bw_file))
```

# Tumor analysis

## Shared mapping & helpers
Group codes in the dataset (e.g. `11636.1`) are translated into clear treatment group names such as *Control* or *Tamoxifen*.  
A fixed color palette is defined to ensure consistent colors across all plots. q2 
```{r mapping}
group_name_map <- c(
  `11636.1` = "Control",
  `11636.2` = "Tamoxifen",
  `11636.3` = "Dexamethasone",
  `11636.4` = "Tamoxifen + Dexamethasone"
)

pal <- c(
  "Control"                   = "#1f77b4",
  "Tamoxifen"                 = "#ff7f0e",
  "Dexamethasone"             = "#2ca02c",
  "Tamoxifen + Dexamethasone" = "#d62728"
)
```

## Tumor: data loading
I imported the tumor measurements per mouse directly from my pre-processed Excel file.  
Before this step, I already took care of two important things myself:  

- **Binning:** I grouped the raw measurements into bins of 1–3 days to reduce day-to-day variability while keeping the overall growth pattern.  
- **Quality check:** I manually reviewed all measurements and marked unreliable values with a `*`. These flagged values are excluded from the dataset before analysis. This also means i took out mice with less than 30 days within the experiment. Also tumor measurements that didnt make sense. 

The resulting dataset `mouse_df` therefore already contains clean, binned tumor volumes with only reliable measurements.  
It includes the following variables:  

- `group_name`: treatment group assignment (Control, Tamoxifen, Dexamethasone, Combination)  
- `mouse_id`: unique identifier per mouse  
- `day`: binned measurement timepoint (e.g. 0, 1–3, 4–6, …)  
- `volume_mm3`: tumor volume in cubic millimeters  

Since I already did the binning and QC, the dataset that is loaded is directly ready for analysis, and is simply sorted by group, mouse, and day. The original raw data is still find able called: `20250623 Catrin Lutz EGP 11636 FINAL iLes export (1).xlsx`


```{r tumor-load}
mouse_df <- readxl::read_excel(tumor_file) %>%
  dplyr::rename_with(tolower) %>%   # make column names lowercase
  dplyr::mutate(
    group_code = as.character(group_code),
    group_name = factor(unname(group_name_map[group_code]), levels = names(pal)),
    day        = as.numeric(day),
    mouse_id   = as.character(mouse_id),
    volume_mm3 = as.numeric(tumor_volume_mm3)
  ) %>%
  dplyr::select(group_code, group_name, mouse_id, day, volume_mm3) %>%
  dplyr::arrange(group_code, mouse_id, day)

knitr::kable(
  head(mouse_df, 20),
  caption = "Preview: tumor per mouse per day",
  format = "markdown"
)

```


## Filtering tumors: remove rows with missing tumor volumes

I first define a filtered dataset that excludes missing tumor volumes ( that i took out ).
This avoids spurious NA rows in tables and plots.
```{r tumor-filter}
mouse_df_filtered <- mouse_df %>%
  filter(!is.na(volume_mm3))

knitr::kable(
  head(mouse_df_filtered, 10),
  caption = "Preview: tumor data after filtering out NA tumor volumes",
  format = "markdown"
)

```




## Tumor growth per group
Tumor growth is visualized for each tumor individually.  
Within each treatment group, a separate plot is generated.  
This allows assessment of variability within groups.   
```{r tumor-individuals, fig.width=12, fig.height=5}
# Plot function for each treatment group separately
plot_group_mice <- function(df_group, title_text) {
  ggplot(df_group, aes(x = day, y = volume_mm3, group = mouse_id, color = mouse_id)) +
    geom_line(linewidth = 1, alpha = 0.9) +
    geom_point(size = 2, alpha = 0.9) +
    scale_color_discrete(guide = "none") +
    labs(
      title = title_text,
      x = "Time (days)",
      y = "Tumor volume (mm³)"
    ) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.major.y = element_line(color = "grey90", linewidth = 0.25),
      panel.grid.minor.y = element_blank()
    )
}

# Loop over treatment groups and make plots
for (g in levels(mouse_df_filtered$group_name)) {
  sub <- filter(mouse_df_filtered, group_name == g)
  
  # Count tumors (L and R separate)
  n_tumors <- n_distinct(sub$mouse_id)
  
  # Count unique mice (ignore L/R suffix)
  n_mice <- n_distinct(gsub("_[LR]$", "", sub$mouse_id))
  
  # Dynamic title
  ttl <- paste0(
    g, " — individual tumor growth (n = ", n_tumors, " tumors; ",
    n_mice, " mice)"
  )
  
  print(plot_group_mice(sub, ttl))
}

```


## Tumor group mean ± SEM (Binned, filtered mice)
To summarize tumor burden across groups, I calculated the mean tumor volume ± SEM per treatment group and day.
This visualization highlights differences between treatments while accounting for variability across tumors.
Only the filtered dataset (without NA tumor volumes) is used here. 

```{r tumor-mean-binned, fig.width=12, fig.height=5}
tumor_daily <- mouse_df_filtered %>%
  group_by(group_name, day) %>%
  summarise(
    mean_volume_mm3 = mean(volume_mm3, na.rm = TRUE),
    sem_mm3         = sd(volume_mm3, na.rm = TRUE) / sqrt(n()),
    n_tumors        = n(),
    .groups = "drop"
  ) %>%
  arrange(group_name, day)

ggplot(tumor_daily, aes(x = day, y = mean_volume_mm3,
                        color = group_name, shape = group_name,
                        group = group_name)) +
  geom_errorbar(aes(ymin = mean_volume_mm3 - sem_mm3,
                    ymax = mean_volume_mm3 + sem_mm3),
                width = 0.3, alpha = 0.9) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_manual(values = pal, breaks = names(pal)) +
  scale_shape_manual(values = c(16, 17, 15, 18), breaks = names(pal)) +  # circle, triangle, square, diamond
  labs(
    title = "Tumor growth by treatment group (mean ± SEM, filtered tumors)",
    x = "Day",
    y = "Tumor volume (mm³)"
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "top",
    legend.title    = element_blank(),   # remove "Treatment"
    legend.text     = element_text(size = 12),
    legend.key.size = unit(1.2, "lines"),  # bigger symbols
    plot.title      = element_text(face = "bold"),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.25),
    panel.grid.minor.y = element_blank()
  )
```



# Bodyweight analysis

Bodyweight analysis serves as an indicator of potential treatment toxicity.  
All mice are included in this analysis (even those excluded in tumor analysis) to avoid bias from selective removal.  

# Bodyweight analysis (unfiltered mice)

## Load bodyweight data
```{r bw-load}
bw_raw <- readxl::read_excel(bw_file, sheet = "Mouse_day_mean")

# Detect correct bodyweight column
bw_candidates <- c("mean_bodyweight_g", "bodyweight_g")
bw_col <- bw_candidates[bw_candidates %in% names(bw_raw)][1]
if (is.na(bw_col)) stop("No bodyweight column found in Mouse_day_mean")

bw_mouse <- bw_raw %>%
  mutate(
    group_code   = as.character(group_code),
    group_name   = factor(unname(group_name_map[group_code]), levels = names(pal)),
    day          = as.numeric(day),
    mouse_id     = as.character(mouse_id),
    bodyweight_g = .data[[bw_col]]
  ) %>%
  filter(group_code %in% names(group_name_map)) %>%
  arrange(group_name, mouse_id, day)

knitr::kable(head(bw_mouse, 8),
             caption="Preview: bodyweight per mouse per day",
             format="markdown")

```

## Bodyweight per mouse
Bodyweight development is shown for each mouse individually.  
Separate plots per treatment group allow for detailed inspection of stability or loss of weight over time.  
```{r bw-plot}
plot_bw_group <- function(df_group, title_text) {
  ggplot(df_group, aes(x = day, y = bodyweight_g,
                       group = mouse_id, color = mouse_id)) +
    geom_line(linewidth = 1, alpha = 0.9) +
    geom_point(size = 2, alpha = 0.9) +
    scale_color_discrete(guide = "none") +
    labs(
      title = title_text,
      x = "Time (days)",
      y = "Bodyweight (g)"
    ) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.major.y = element_line(color = "grey90", linewidth = 0.25),
      panel.grid.minor.y = element_blank()
    )
}

for (g in levels(bw_mouse$group_name)) {
  sub <- filter(bw_mouse, group_name == g)
  n_m <- n_distinct(sub$mouse_id)
  ttl <- paste0(g, " — individual bodyweights (n = ", n_m, ")")
  print(plot_bw_group(sub, ttl))
}
```

## Bodyweight group mean ± SEM
To summarize bodyweight development across groups, I calculated the mean bodyweight ± SEM per treatment group and day.
To avoid bias from less measurments in a day, I excluded all days with fewer than 7 individual mouse measurements.
This ensures that the plotted averages are based on sufficient observations and reflect reliable group-level trends. 
```{r,bw-mean,SEM}

bw_daily <- bw_mouse %>%
  group_by(group_name, day) %>%
  summarise(
    mean_bw = mean(bodyweight_g, na.rm = TRUE),
    sem_bw  = sd(bodyweight_g, na.rm = TRUE) / sqrt(n()),  # SEM
    n       = n(),
    .groups = "drop"
  ) %>%
  filter(n >= 7) %>%   # keep only days with ≥ 7 mice measured
  arrange(group_name, day)

ggplot(bw_daily, aes(x = day, y = mean_bw,
                     color = group_name, group = group_name, shape = group_name)) +
  geom_errorbar(aes(ymin = mean_bw - sem_bw, ymax = mean_bw + sem_bw),
                width = 0.3, alpha = 0.9) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.6) +
  scale_color_manual(values = pal, breaks = names(pal), name = "Treatment") +
  scale_shape_manual(values = c(16, 17, 15, 18), name = "Treatment") +
  labs(
    title = "Bodyweight development by treatment group (mean ± SEM, days with ≥ 7 mice)",
    x = "Day",
    y = "Bodyweight (g)"
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "top",
    legend.title    = element_text(size = 11, face = "bold"),
    legend.text     = element_text(size = 10),
    plot.title      = element_text(face = "bold"),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.25),
    panel.grid.minor.y = element_blank()
  )
```


## Bodyweight change before and after treatment break
To assess the impact of the treatment break on bodyweight, I compared the percentage change from baseline bodyweight before the break (days 0–27) and after recovery (days ≥36).
For each mouse, the relative change compared to its baseline (day 0) was calculated, and then averaged within treatment groups.
The resulting bar plot shows the mean bodyweight change ± SEM per group, highlighting both the weight loss observed during treatment and the partial recovery after treatment was paused.
The number of mice contributing to each estimate is shown above the bars.
```{r bw-barplot, fig.width=12, fig.height=5}
## Bodyweight change before vs after treatment break (same mice only)

# Calculate % change from baseline
baseline_bw <- bw_mouse %>%
  group_by(mouse_id) %>%
  summarise(baseline = first(bodyweight_g), .groups = "drop")

bw_pct <- bw_mouse %>%
  left_join(baseline_bw, by = "mouse_id") %>%
  mutate(pct_change = 100 * (bodyweight_g - baseline) / baseline)

# Restrict to mice with measurements in both periods (before 28, after 36)
valid_mice <- bw_pct %>%
  group_by(mouse_id) %>%
  summarise(
    has_before = any(day < 28, na.rm = TRUE),
    has_after  = any(day >= 36, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(has_before & has_after) %>%
  pull(mouse_id)

bw_pct_complete <- bw_pct %>%
  filter(mouse_id %in% valid_mice)

# Summarize means for before and after (exclude break window 28–35)
bw_summary <- bw_pct_complete %>%
  mutate(period = case_when(
    day < 28 ~ "Before day 28",
    day >= 36 ~ "After day 36",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(period)) %>%
  group_by(group_name, period, mouse_id) %>%
  summarise(mean_change = mean(pct_change, na.rm = TRUE), .groups = "drop") %>%
  group_by(group_name, period) %>%
  summarise(
    mean_change = mean(mean_change, na.rm = TRUE),
    sem_change  = sd(mean_change, na.rm = TRUE) / sqrt(n()),
    n_mice      = n(),
    .groups = "drop"
  ) %>%
  mutate(period = factor(period, levels = c("Before day 28", "After day 36")))

# Plot with baseline at 0% and normal % labels
ggplot(bw_summary, aes(x = period, y = mean_change, fill = group_name)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7, alpha = 0.9) +
  geom_errorbar(aes(ymin = mean_change - sem_change,
                    ymax = mean_change + sem_change),
                width = 0.2,
                position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", mean_change)),
            position = position_dodge(width = 0.8),
            vjust = ifelse(bw_summary$mean_change >= 0, -0.5, 1.2),
            size = 4) +
  scale_fill_manual(values = pal, name = "Treatment") +
  labs(
    title = "Mean bodyweight change before vs after treatment break (same mice only)",
    x = "",
    y = "% change from baseline (mean ± SEM)"
  ) +
  expand_limits(y = c(-10, 7)) +   # extend Y-axis so labels aren’t cut off
  theme_classic(base_size = 13) +
  theme(
    legend.position = "top",
    legend.title    = element_text(size = 11, face = "bold"),
    legend.text     = element_text(size = 10),
    plot.title      = element_text(face = "bold", size = 14),
    axis.text.x     = element_text(angle = 20, hjust = 1)
  )


```



## Statistical comparison: Before vs After 

To formally test whether bodyweight significantly increased after the treatment break compared to before, we performed within-mouse paired comparisons.
For each mouse, the mean percentage bodyweight change (relative to baseline) was calculated separately for the periods before day 28 and after day 36.
These paired values were then compared within each treatment group using:

Paired t-test: assumes approximately normally distributed differences.

Wilcoxon signed-rank test: a non-parametric alternative that does not assume normality.

The resulting table reports the mean difference (After−Before, %), number of mice, and p-values from both tests.

```{r,Statistical comparison}
# --- Statistical comparison of bodyweight change before vs after treatment break ---

# Calculate % change from baseline per mouse
baseline_bw <- bw_mouse %>%
  group_by(mouse_id) %>%
  summarise(baseline = first(bodyweight_g), .groups = "drop")

bw_pct <- bw_mouse %>%
  left_join(baseline_bw, by = "mouse_id") %>%
  mutate(pct_change = 100 * (bodyweight_g - baseline) / baseline)

# Restrict to mice with both before (day <28) and after (day ≥36)
valid_mice <- bw_pct %>%
  group_by(mouse_id) %>%
  summarise(
    has_before = any(day < 28, na.rm = TRUE),
    has_after  = any(day >= 36, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(has_before & has_after) %>%
  pull(mouse_id)

bw_pct_complete <- bw_pct %>%
  filter(mouse_id %in% valid_mice)

# Summarize per mouse: mean before vs after
bw_mouse_summary <- bw_pct_complete %>%
  mutate(period = case_when(
    day < 28 ~ "Before",
    day >= 36 ~ "After",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(period)) %>%
  group_by(group_name, mouse_id, period) %>%
  summarise(mean_change = mean(pct_change, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = period, values_from = mean_change)

# Stats per group: paired t-test + Wilcoxon
stats_table <- bw_mouse_summary %>%
  group_by(group_name) %>%
  summarise(
    N        = n(),
    mean_diff = mean(After - Before, na.rm = TRUE),
    ttest_p   = t.test(After, Before, paired = TRUE)$p.value,
    wilcox_p  = wilcox.test(After, Before, paired = TRUE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    mean_diff = round(mean_diff, 2),
    ttest_p   = ifelse(ttest_p < 0.0001, "<0.0001",
                       formatC(ttest_p, format = "f", digits = 4)),
    wilcox_p  = ifelse(wilcox_p < 0.0001, "<0.0001",
                       formatC(wilcox_p, format = "f", digits = 4))
  )

# Show as markdown table
knitr::kable(
  stats_table,
  col.names = c("Treatment", "N", "Mean Δ (After−Before, %)", 
                "Paired t-test p", "Wilcoxon p"),
  align = "lcccc",
  caption = "Statistical comparison of bodyweight change before vs after treatment break"
)
```
```{r}


# Filter dataset for the two groups of interest
df_sub <- mouse_df_filtered %>%
  filter(group_name %in% c("Tamoxifen", "Tamoxifen + Dexamethasone"))

results <- list()

# Loop over real day numbers
for (d in sort(unique(df_sub$day))) {
  sub <- df_sub %>% filter(day == d)
  
  if (n_distinct(sub$group_name) == 2) {
    fit <- lm(volume_mm3 ~ group_name, data = sub)
    emm <- emmeans(fit, ~ group_name)
    tukey <- pairs(emm, adjust = "tukey") %>% as.data.frame()
    
    row <- tukey %>%
      filter(grepl("Tamoxifen", contrast) & grepl("Dexamethasone", contrast)) %>%
      mutate(day = d,  # <-- keep the *real* numeric day
             mean_diff = estimate,
             p_adj = p.value,
             reject = p.value < 0.05) %>%
      select(day, mean_diff, p_adj, reject)
    
    results[[length(results) + 1]] <- row
  }
}

# Final results with *true day values*
results_df <- bind_rows(results) %>% arrange(day)

knitr::kable(results_df, digits = 4,
             caption = "Tamoxifen vs Tamoxifen + Dexamethasone per day (Tukey-adjusted)")


```



```{r}
library(dplyr)
library(emmeans)
library(knitr)

# Filter dataset for the two groups of interest
df_sub <- mouse_df_filtered %>%
  filter(group_name %in% c("Tamoxifen", "Tamoxifen + Dexamethasone"))

results <- list()

# Loop over all measurement days
for (d in sort(unique(df_sub$day))) {
  sub <- df_sub %>% filter(day == d)
  
  if (n_distinct(sub$group_name) == 2) {
    # Fit linear model for that day
    fit <- lm(volume_mm3 ~ group_name, data = sub)
    emm <- emmeans(fit, ~ group_name)
    tukey <- pairs(emm, adjust = "tukey") %>% as.data.frame()
    
    # Extract Tamoxifen vs Tamoxifen+Dexa
    row <- tukey %>%
      filter(grepl("Tamoxifen", contrast) & grepl("Dexamethasone", contrast)) %>%
      mutate(day = d,
             mean_diff = estimate,
             p_adj = p.value,
             reject = p.value < 0.05) %>%
      select(day, mean_diff, p_adj,)
    
    results[[length(results) + 1]] <- row
  }
}

# Collect results
results_df <- bind_rows(results) %>% arrange(day)

# Show only Day 60
day60_result <- results_df %>% filter(day == 60)

kable(day60_result, digits = 4,
      caption = "Tamoxifen vs Tamoxifen + Dexamethasone at Day 60 (Tukey-adjusted)")

```




